sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/model-description.Rmd
Preping JSS

---
title: "Appendix 1: sdmTMB statistical model description"
output:
  bookdown::pdf_document2:
    highlight: pygments
    toc: true
    number_sections: true
fig_caption: true
urlcolor: blue
highlight: "monochrome"
csl: mee.csl
bibliography: refs.bib
header-includes:
  \usepackage[left]{lineno}
  \usepackage{bm}
  \usepackage{amssymb}
  \linenumbers
---

```{r setup, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
```

@@

# Notation conventions

@@

```{r indices}
library(magrittr)
subscripts <- tibble::tribble(
  ~Symbol, ~Description,
  "$\\s$", "Index for space; a vector of x and y coordinates",
  "$t$", "Index for time",
  "$g$", "Group"
)
knitr::kable(subscripts, caption = "Subscript notation", escape = FALSE, format = "latex", booktabs = TRUE) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

```{r symbols, results='asis'}
symbols <- tibble::tribble(
  ~Symbol, ~Code, ~Description,
  "$y$", "\\texttt{y\\_i}", "Observed response data",
  "$\\mu$", "\\texttt{mu\\_i}", "Mean",
  "$\\phi$", "\\texttt{phi}", "A dispersion parameter for a distribution",
  "$f$", "\\texttt{fit\\$family\\$link}", "Link function",
  "$f^{-1}$", "\\texttt{fit\\$family\\$linkinv}", "Inverse link function",
  "$\\boldsymbol{\\beta}$", "\\texttt{b\\_j}", "Parameter vector",
  "$\\boldsymbol{X}$", "\\texttt{X\\_ij}", "A predictor model matrix",
  "$O_{\\s, t}$", "\\texttt{offset}", "An offset variable at point $\\s$ and time $t$",

  "$\\omega_{\\s}$", "\\texttt{omega\\_s}", "Spatial random field at point $\\s$ (knot)",
  "$\\omega_{\\s}^*$", "\\texttt{omega\\_s\\_A}", "Spatial random field at point $\\s$ (interpolated)",
  "$\\zeta_{\\s}$", "\\texttt{zeta\\_s}", "Spatially varying coefficient random field at point $\\s$ (knot)",
  "$\\zeta_{\\s}^*$", "\\texttt{zeta\\_s\\_A}", "Spatially varying coefficient random field at point $\\s$ (interpolated)",
  "$\\epsilon_{\\s, t}$", "\\texttt{epsilon\\_st}", "Spatiotemporal random field at point $\\s$ and time $t$ (knot)",
  "$\\epsilon_{\\s, t}^*$", "\\texttt{epsilon\\_st\\_A}", "Spatiotemporal random field at point $\\s$ and time $t$ (interpolated)",
  "$\\delta_{\\s,t}$", "\\texttt{b\\_t}", "AR(1) or random walk spatiotemporal deviations (knot)",
  "$\\alpha_g$", "\\texttt{RE}", "IID random intercept deviation for group $g$",

  "$\\boldsymbol{\\Sigma}_\\omega$", "\\texttt{-}", "Spatial random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\zeta$", "\\texttt{-}", "Spatially varying coefficient random field covariance matrix",
  "$\\boldsymbol{\\Sigma}_\\epsilon$", "\\texttt{-}", "Spatiotemporal random field covariance matrix",
  "$\\boldsymbol{Q}_\\omega$", "\\texttt{Q\\_s}", "Spatial random field precision matrix",
  "$\\boldsymbol{Q}_\\zeta$", "\\texttt{Q\\_s}", "Spatially varying coefficient random field precision matrix",
  "$\\boldsymbol{Q}_\\epsilon$", "\\texttt{Q\\_st}", "Spatiotemporal random field precision matrix",

  "$\\sigma_\\alpha^2$", "\\texttt{sigma\\_G}", "IID random intercept variance",
  "$\\sigma_\\epsilon^2$", "\\texttt{sigma\\_E}", "Spatiotemporal random field marginal variance",
  "$\\sigma_\\omega^2$", "\\texttt{sigma\\_O}", "Spatial random field marginal variance",
  "$\\sigma_\\zeta^2$", "\\texttt{sigma\\_Z}", "Spatially varying coefficient random field marginal variance",
  "$\\kappa_\\omega$", "\\texttt{kappa(0)}", "Spatial decorrelation rate",
  "$\\kappa_\\epsilon$", "\\texttt{kappa(1)}", "Spatiotemporal decorrelation rate",

  "$\\rho$", "\\texttt{ar1\\_rho}", "Correlation between random fields in subsequent time steps",
  "$\\boldsymbol{A}$", "\\texttt{A}", "Sparse projection matrix to interpolate between knot and data locations",
  "$\\boldsymbol{H}$", "\\texttt{H}", "2-parameter rotation matrix used to define anisotropy"
)
knitr::kable(symbols, caption = "Symbol notation, code representation (in model output or in model template code), and descriptions.", escape = FALSE, format = "latex", booktabs = TRUE,
   linesep = c(
    rep('', 7), # y, mean etc.
    '\\addlinespace',
    rep('', 7), # fields
    '\\addlinespace',
    rep('', 5), # covariance stuff
    '\\addlinespace',
    rep('', 5), # more covariance stuff
    '\\addlinespace',
    rep('', 99) # end
  ))
```

@@

# sdmTMB model structure

@@

## Main effects

@@

### Smoothers

@@

### Linear break-point threshold models

@@

### Logistic threshold models

@@

## Spatial random fields

@@

## Spatiotemporal random fields

@@

### IID spatiotemporal random fields

@@

### AR(1) spatiotemporal random fields

@@

### Random walk spatiotemporal random fields (RW)

@@

## Time-varying regression parameters

@@

## Spatially varying coefficients (SVC)

@@

## IID random or multi-level intercepts

@@

## Offset terms

@@

# Observation model families

@@

## Binomial

@@

## Beta

@@

## Gamma

@@

## Gaussian

@@

## Lognormal

@@

Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the TMB `dnorm()` normal density.

@@

## Negative Binomial 1 (NB1)

@@

## Negative Binomial 2 (NB2)

@@

## Poisson

@@

## Student-t

@@

## Tweedie

@@

# Gaussian random fields

## Matérn parameterization

@@

## Projection $\boldsymbol{A}$ matrix

@@

## Anisotropy

@@

## Incorporating physical barriers into the SPDE

@@

# Optimization

## Optimization details

@@

## Assessing convergence

@@

# References

>>>>

# Model

@@

## Notation conventions

@@

\begin{table}[!h]
\caption{Subscript notation}
\centering
\begin{tabular}[t]{ll}
\toprule
Symbol & Description\\
\midrule
$\s$ & Index for space; a vector of x and y coordinates\\
$t$ & Index for time\\
$g$ & Group\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\caption{Symbol notation, code representation (in model output or in model template code), and descriptions.}
\centering
\begin{tabular}[t]{lll}
\toprule
Symbol & Code & Description\\
\midrule
$y$ & \texttt{y\_i} & Observed response data\\
$\mu$ & \texttt{mu\_i} & Mean\\
$\phi$ & \texttt{phi} & A dispersion parameter for a distribution\\
$f$ & \texttt{fit\$family\$link} & Link function\\
$f^{-1}$ & \texttt{fit\$family\$linkinv} & Inverse link function\\
$\boldsymbol{\beta}$ & \texttt{b\_j} & Parameter vector\\
$\boldsymbol{X}$ & \texttt{X\_ij} & A predictor model matrix\\
$O_{\s, t}$ & \texttt{offset} & An offset variable at point $\s$ and time $t$\\
\addlinespace
$\omega_{\s}$ & \texttt{omega\_s} & Spatial random field at point $\s$ (knot)\\
$\omega_{\s}^*$ & \texttt{omega\_s\_A} & Spatial random field at point $\s$ (interpolated)\\
$\zeta_{\s}$ & \texttt{zeta\_s} & Spatially varying coefficient random field at point $\s$ (knot)\\
$\zeta_{\s}^*$ & \texttt{zeta\_s\_A} & Spatially varying coefficient random field at point $\s$ (interpolated)\\
$\epsilon_{\s, t}$ & \texttt{epsilon\_st} & Spatiotemporal random field at point $\s$ and time $t$ (knot)\\
$\epsilon_{\s, t}^*$ & \texttt{epsilon\_st\_A} & Spatiotemporal random field at point $\s$ and time $t$ (interpolated)\\
$\delta_{\s,t}$ & \texttt{b\_t} & AR(1) or random walk spatiotemporal deviations (knot)\\
$\alpha_g$ & \texttt{RE} & IID random intercept deviation for group $g$\\
\addlinespace
$\boldsymbol{\Sigma}_\omega$ & \texttt{-} & Spatial random field covariance matrix\\
$\boldsymbol{\Sigma}_\zeta$ & \texttt{-} & Spatially varying coefficient random field covariance matrix\\
$\boldsymbol{\Sigma}_\epsilon$ & \texttt{-} & Spatiotemporal random field covariance matrix\\
$\boldsymbol{Q}_\omega$ & \texttt{Q\_s} & Spatial random field precision matrix\\
$\boldsymbol{Q}_\zeta$ & \texttt{Q\_s} & Spatially varying coefficient random field precision matrix\\
$\boldsymbol{Q}_\epsilon$ & \texttt{Q\_st} & Spatiotemporal random field precision matrix\\
\addlinespace
$\sigma_\alpha^2$ & \texttt{sigma\_G} & IID random intercept variance\\
$\sigma_\epsilon^2$ & \texttt{sigma\_E} & Spatiotemporal random field marginal variance\\
$\sigma_\omega^2$ & \texttt{sigma\_O} & Spatial random field marginal variance\\
$\sigma_\zeta^2$ & \texttt{sigma\_Z} & Spatially varying coefficient random field marginal variance\\
$\kappa_\omega$ & \texttt{kappa(0)} & Spatial decorrelation rate\\
$\kappa_\epsilon$ & \texttt{kappa(1)} & Spatiotemporal decorrelation rate\\
\addlinespace
$\rho$ & \texttt{ar1\_rho} & Correlation between random fields in subsequent time steps\\
$\boldsymbol{A}$ & \texttt{A} & Sparse projection matrix to interpolate between knot and data locations\\
$\boldsymbol{H}$ & \texttt{H} & 2-parameter rotation matrix used to define anisotropy\\
\bottomrule
\end{tabular}
\end{table}

@@

## sdmTMB model structure

@@

### Main effects

@@

#### Smoothers

@@

#### Linear break-point threshold models

@@

#### Logistic threshold models

@@

### Spatial random fields

@@

### Spatiotemporal random fields

@@

#### IID spatiotemporal random fields

@@

#### AR(1) spatiotemporal random fields

@@

#### Random walk spatiotemporal random fields (RW)

@@

### Time-varying regression parameters

@@

### Spatially varying coefficients (SVC)

@@

### IID random or multi-level intercepts

@@

### Offset terms

@@

## Observation model families

@@

### Binomial

@@

### Beta

@@

### Gamma

@@

### Gaussian

@@

### Lognormal

@@

Code defined [within sdmTMB](https://github.com/pbs-assess/sdmTMB/blob/18a39eabc111e2179fa589f942c8820d87ad10df/src/utils.h#L47-L54) based on the \pkg{TMB} `dnorm()` normal density.

@@

### Negative Binomial 1 (NB1)

@@

### Negative Binomial 2 (NB2)

@@

### Poisson

@@

### Student-t

@@

### Tweedie

@@

## Gaussian random fields

### Matérn parameterization

@@

### Projection A matrix

@@

### Anisotropy

@@

### Incorporating physical barriers into the SPDE

@@

## Optimization

### Optimization details

@@

### Assessing convergence

@@
